---
output: github_document
editor_options: 
  chunk_output_type: console
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# advent-of-code
Attempting the 2021 [advent of code](https://adventofcode.com/) in **base R** only!

I am a massive fan and regular user of the tidyverse, but every now and then I like to remind myself of what life used to be like...

# --- Day 1: Sonar Sweep ---
## 1a
```{r}
d <- as.numeric(readLines('data/1-input.txt'))
sum(diff(d) > 0)
```

## 1b
```{r}
r <- sapply(1:(length(d)-length(d)%%3), function(x) sum(d[x:(x+2)]))
sum(diff(r) > 0)
```

# --- Day 2: Dive! ---
## 2a
```{r}
d <- readLines('data/2-input.txt')
direction <- sapply(strsplit(d, " "), `[[`, 1)
magnitude <- as.numeric(sapply(strsplit(d, " "), `[[`, 2))

horizontal <- sum(magnitude[direction == "forward"])
depth <- sum(magnitude[direction == "down"]) + sum(-magnitude[direction == "up"])
horizontal * depth
```

## 2b
```{r}
l <- length(d) + 1
aim <- rep(0, l)
depth <- rep(0, l)
horiz <- rep(0, l)

for(i in seq_along(d)){
  if(direction[i] == "down"){
    aim[i+1] <- aim[i] + magnitude[i]
    depth[i+1] <- depth[i]
    horiz[i+1] <- horiz[i]
  } else if(direction[i] == "up"){
    aim[i+1] <- aim[i] - magnitude[i]
    depth[i+1] <- depth[i]
    horiz[i+1] <- horiz[i]
  } else if(direction[i] == "forward"){
    horiz[i+1] <- horiz[i] + magnitude[i]
    depth[i+1] <- depth[i] + (aim[i] * magnitude[i])
    aim[i+1] <- aim[i]
  }
}

horiz[l] * depth[l]
```

# --- Day 3: Binary Diagnostic ---
## 3a
```{r}
d <- readLines('data/3-input.txt')
m <- matrix(as.integer(unlist(strsplit(d, ""))), ncol = 12, byrow = TRUE)

gamma <- strtoi(paste0(as.integer(colMeans(m) > 0.5), collapse=""), base=2)
epsilon <- strtoi(paste0(as.integer(colMeans(m) < 0.5), collapse=""), base=2)

gamma * epsilon
```

## 3b
```{r}
# O2
mi <- m
for(i in 1:ncol(mi)){
  if(nrow(mi) == 1) break
  mi <- mi[which(mi[,i] == round(ifelse(mean(mi[,i]) == 0.5, 1, mean(mi[,i])))),,drop = F]}
o2 <- strtoi(paste0(as.vector(mi), collapse = ""), base=2)

# CO2
mi <- m
for(i in 1:ncol(mi)){
  if(nrow(mi) == 1) break
  mi <- mi[which(mi[,i] == !round(ifelse(mean(mi[,i]) == 0.5, 1, mean(mi[,i])))),,drop = F]}
co2 <- strtoi(paste0(as.vector(mi), collapse = ""), base=2)

o2 * co2
```


# --- Day 4: Giant Squid ---
## 4a
```{r}
d <- readLines('data/4-input.txt')

# Extract numbers
nums <- as.integer(unlist(strsplit(d[1], ",")))

# Extract bingo boards
# trimws() to remove leading spaces for single digit numbers
# gsub() all double spaces to single spaces so I can strsplit() on single spaces
boards <-
  lapply(
    lapply(which(d == ""), function(x) gsub("  ", " ", trimws(d[(x+1):(x+5)]))),
    function(x) matrix(as.integer(unlist(strsplit(paste0(x, combine=" "), " "))),
                       ncol=5, byrow=TRUE))
```

* Function to test if a matrix has a completed column or row (where completed numbers are NA)
```{r}
is_bingo <- function(m){
  
  check_rows <- apply(m, MARGIN = 1, function(x) all(is.na(x)))
  check_cols <- apply(m, MARGIN = 2, function(x) all(is.na(x)))
  
  if(any(check_cols) | any(check_rows)){return(TRUE)} else {return(FALSE)}
}
```

```{r}
boards_tmp <- boards

for(i in seq_along(nums)){
  
  # Update boards
  boards_tmp <- lapply(boards_tmp, function(x) replace(x, x==nums[i], values = NA))
  
  # Test for any with BINGO
  results <- sapply(boards_tmp, is_bingo)

  # If any board has BINGO, return results
  if(any(results)){
    print(paste0("BINGO!"))
    # hopefully only one board wins per number called...
    winning_board <- boards_tmp[[which(results)[1]]] 
    print(winning_board)
    winning_board_sum <- sum(winning_board[!is.na(winning_board)])
    number_called <- nums[i]
    break}
  }

winning_board_sum * number_called
```

## 4b
```{r}
boards_tmp <- boards

for(i in seq_along(nums)){
  
  # Update boards
  boards_tmp <- lapply(boards_tmp, function(x) replace(x, x==nums[i], values = NA))
  
  # Test for any with BINGO
  results <- sapply(boards_tmp, is_bingo)
  
  # If only one board remaining to get BINGO - save the ID of the board
  # This will then run every time until the final board is solved
  if(sum(!results) == 1){
    print("One board left to get BINGO")
    last_board_id <- which(!results)
  }

  # When all boards are solved, return the final board (solved) and the final
  # number called
  if(all(results)){
    print("All boards have BINGO")
    final_board <- boards_tmp[[last_board_id]]
    final_number <- nums[i]
    break
  }
}

sum(final_board[!is.na(final_board)]) * final_number
```

# --- Day 5: Hydrothermal Venture ---
## 5a
* Read data and wrangle to dataframe
```{r}
d <- 
  do.call(rbind,
  lapply(
    strsplit(readLines('data/5-input.txt'), ' -> '),
    function(x) data.frame(x1 = as.numeric(strsplit(x, ",")[[1]][1]),
                           y1 = as.numeric(strsplit(x, ",")[[1]][2]),
                           x2 = as.numeric(strsplit(x, ",")[[2]][1]),
                           y2 = as.numeric(strsplit(x, ",")[[2]][2]))))
```

* Visaulise all lines for interest
```{r fig.width=6, fig.height=6}
xmin <- min(c(d$x1, d$x2))
xmax <- max(c(d$x1, d$x2))
ymin <- min(c(d$y1, d$y2))
ymax <- max(c(d$y1, d$y2))

plot.new()
plot.window(xlim = c(xmin, xmax), ylim = c(ymin, ymax), asp=1)
segments(x0 = d$x1, x1 = d$x2, y0 = d$y1, y1 = d$y2)
axis(1)
axis(2)
```

* Subset to pure horizontal or vertical lines
* Initialise empty list for output
* Generate all coordinates along each line
```{r}
sl <- d[(d$x1 == d$x2) | (d$y1 == d$y2),]

l <- list()
for(i in 1:nrow(sl)){
  if(sl$x1[i] == sl$x2[i]){
    # Pure horizontal
    l[[i]] <- paste0(sl$x1[i], ",", (sl$y1[i]:sl$y2[i]))
  } else if(sl$y1[i] == sl$y2[i]){
    # Pure vertical
    l[[i]] <- paste0((sl$x1[i]:sl$x2[i]), ",",  sl$y1[i])
  }
}

# Count unique coords and how many are appear twice or more
counts <- table(unlist(l))
length(counts[counts >=2])
```

## 5b
* Same as above but on the full data set adding a third else statement for diagonal lines
* The question tells us diagonals are only ever at 45 degrees - so the computation of the coordinates is simple
```{r}
l <- list()
for(i in 1:nrow(d)){
  if(d$x1[i] == d$x2[i]){
    # Pure horizontal
    l[[i]] <- paste0(d$x1[i], ",", (d$y1[i]:d$y2[i]))
  } else if(d$y1[i] == d$y2[i]){
    # Pure vertical
    l[[i]] <- paste0((d$x1[i]:d$x2[i]), ",",  d$y1[i])
  } else {
    # Diagonal
    l[[i]] <- paste0((d$x1[i]:d$x2[i]), ",",  (d$y1[i]:d$y2[i]))
  }
}

counts <- table(unlist(l))
length(counts[counts >=2])
```

# --- Day 6: Lanternfish ---
> Each day, a 0 becomes a 6 and adds a new 8 to the end of the list, while each other number decreases by 1 if it was present at the start of the day.

* Read data
```{r}
d <- scan('data/6-input.txt', integer(), sep = ",")
```

## 6a
* Create a dataframe showing the count of each day value 
* Left join to all possible day values (0:8) so I have all options
* Iterate over 1 to 80 days shifting the day values by 1
  * Zeros roll over to be eights (newly born lantern fish)
  * Zeros also needed to be added to the current number of sixes as they start their cycle again
```{r}
x <- table(d)
o <- merge(data.frame(d = 0:8), data.frame(x), all = TRUE)
o$Freq[is.na(o$Freq)] <- 0

for(i in 1:80){
  births <- o$Freq[o$d == 0]
  o$d <- c(o$d[9], o$d[1:8])
  o$Freq[o$d == 6] <- o$Freq[o$d == 6] + births}

sum(o$Freq)
```

## 6b
* Run again for 256 days
```{r}
x <- table(d)
o <- merge(data.frame(d = 0:8), data.frame(x), all = TRUE)
o$Freq[is.na(o$Freq)] <- 0

for(i in 1:256){
  births <- o$Freq[o$d == 0]
  o$d <- c(o$d[9], o$d[1:8])
  o$Freq[o$d == 6] <- o$Freq[o$d == 6] + births}

options(scipen = 999)
sum(o$Freq)
```

* For my own interest - plot the exponential growth and model it
```{r fig.width=6, fig.height=6}
x <- table(d)
o <- merge(data.frame(d = 0:8), data.frame(x), all = TRUE)
o$Freq[is.na(o$Freq)] <- 0

df <- data.frame(day = 1:40, n=NA)

for(i in 1:nrow(df)){
  births <- o$Freq[o$d == 0]
  o$d <- c(o$d[9], o$d[1:8])
  o$Freq[o$d == 6] <- o$Freq[o$d == 6] + births
  df$n[df$day == i] <- sum(o$Freq)}

mod <- nls(n~a*exp(b*day), data = df, start = list(a=1, b=0.1))
summary(mod)
plot(df$day, df$n)
lines(predict(mod), col=2)
```

